### 2017-07-04 abstract.tex

  To address this issue, we present a novel program logic that enables
  compositional reasoning about the behavior of concurrently executing
  weakly-isolated transactions.  Recognizing that the proof burden
  necessary to use this logic may dissuade application developers, we
  also describe a fully automated verification tool based on this
  foundation that determines the safety of high-level consistency
  assertions associated with such transactions.  The key to effective
  inference is the observation that weakly-isolated transactions can
  be viewed as functional (monadic) computations over database state,
  allowing us to treat their operations as state transformers over the
  database.  This interpretation enables automated verification using
  off-the-shelf SMT solvers.

  Notably, our development is parametric over a transaction's specific
  isolation semantics, allowing it to be applicable over a range of
  concurrency control mechanisms.  Case studies and experiments on
  real-world applications demonstrate the utility of our approach, and
  provide strong evidence that automated verification of
  weakly-isolated transactions can be placed on the same formal
  footing as their strongly-isolated serializable counterparts.

>>>>

  To address this issue, we present a novel program logic that enables
  compositional reasoning about the behavior of concurrently executing
  weakly-isolated transactions.  Recognizing that the proof burden
  necessary to use this logic may dissuade application developers, we
  also describe an inference procedure based on this foundation that
  ascertains the weakest isolation level that still guarantees the
  safety of high-level consistency assertions associated with such
  transactions.  The key to effective inference is the observation
  that weakly-isolated transactions can be viewed as functional
  (monadic) computations over database state, allowing us to treat
  their operations as state transformers over the database.  This
  interpretation enables automated verification using off-the-shelf
  SMT solvers.

  Notably, our development is parametric over a transaction's specific
  isolation semantics, allowing it to be applicable over a range of
  concurrency control mechanisms.  Case studies and experiments on
  real-world applications (written in an embedded DSL in OCaml)
  demonstrate the utility of our approach, and provide strong evidence
  that automated verification of weakly-isolated transactions can be
  placed on the same formal footing as their strongly-isolated
  serializable counterparts.

### 2017-07-04 introduction.tex

Unfortunately, weak isolation admits behaviors that are difficult to
comprehend~\cite{berenson}. To quantify weak isolation anomalies,
Fekete \emph{et al.}~\cite{feketevldb09} devised and experimented with
a microbenchmark suite that executes transactions under a
weakly-isolated \emph{read committed} isolation level - the default
level for 8 of the 18 databases studied in~\cite{bailishotos}, and
found that 25 out of every 1000 rows in the database violate at least
one integrity constraint. Bailis \emph{et al.}~\cite{bailisferal} rely
on Rails' \emph{uniqueness validation} to maintain uniqueness of
records while serving Linkbench's~\cite{linkbench} insertion workload
(6400 records distributed over 1000 keys; 64 concurrent clients), and
report discovering more than 10 duplicate records.  Rails relies on
database transactions to validate uniqueness during insertions, which
is sensible if transactions are serializable, but incorrect under the
\emph{read committed} isolation level used in the experiments. The
same study has found that 13\% of all invariants among 67 open source
Ruby-on-Rails applications are at risk of being violated due to weak
isolation. Indeed, incidents of safety violations due to executing
applications in a weakly-isolated environment have been reported on
web services in production~\cite{starbucksbug, scimedbug}, including
in safety-critical applications such as bitcoin
exchanges~\cite{poloniexbug, bitcoinbug}. While enforcing
serializability for all transactions would be sufficient to avoid
these errors and anomalies, it would likely be an overly conservative
strategy; indeed, 75\% of the invariants studied in~\cite{bailisferal}
were shown to be preserved under some form of weak isolation.  When to
use weak isolation, and in what form, is therefore a prominent
question facing all database programmers.\footnote{This position has
  been echoed by leading database researchers who lament the lack of
  better understanding of this problem.  See: {\tt
    http://www.bailis.org/blog/understanding-weak-isolation-is-a-serious-problem}.}

@@

In this paper, we propose a program logic for weakly-isolated
transactions along with automated verification support that realizes
this goal.  We develop a set of syntax-directed compositional proof
rules that allow developers to build correctness proofs for
transactional programs in the presence of a weakly-isolated
concurrency control mechanism.  Moreover, we also present an inference
algorithm based on a state transformer semantics that allows us to
automatically verify the weakest isolation level for a transaction
that ensures its consistency requirements are maintained.  The key to
inference is our formulation of database state (represented as sets of
tuples) as a monad, in which database computations are interpreted as
transformers over these sets.  This interpretation leads to an
encoding database operations amenable for verification by SMT solvers.

We have realized these ideas as an embedded DSL within OCaml, with
support for common SQL operations (updates, selects, inserts, deletes,
joins, etc.).  Experimental results on real-world benchmarks
demonstrate the feasibility of automated verification of weakly
isolated transactions, an important advance in improving the safety of
realistic database computations. The paper makes the following
contributions:
\vspace*{-6pt}
\begin{enumerate}
\item An operational semantics for an embedded DSL in OCaml that
  expresses SQL-based relational database operations as monadic
  computations on database state represented abstractly as sets of
  tuples.
 \item A compositional proof system for this language capable of
   relating high-level application invariants to transformations on
   database state, parameterized by a weak isolation semantics that
   selectively exposes the visibility of these operations to other
   transactions.
 \item An inference algorithm capable of discovering the
   weakest isolation level that is sound with respect to a
   transaction's high-level consistency requirements.  The algorithm
   exploits our monadic formulation, treating database computations as
   predicate state transformers, that leads to a natural encoding
   amenable for automated verification.
 \item A realistic implementation along with results on real-world
   database benchmarks that justify our approach and the utility of
   our proof methodology.
\end{enumerate}
\noindent Our results provide the first (to the best of our knowledge)
automated mechanism to verify the safety of high-level consistency
conditions for weakly-isolated transactions, thereby allowing such
transactions to enjoy the same rigorous reasoning capabilities as
their strongly-isolated (serializable) counterparts.

>>>>

Unfortunately, weak isolation admits behaviors that are difficult to
comprehend~\cite{berenson}. To quantify weak isolation anomalies,
Fekete \emph{et al.}~\cite{feketevldb09} devised and experimented with
a microbenchmark suite that executes transactions under a
weakly-isolated \emph{read committed} isolation level - the default
level for 8 of the 18 databases studied in~\cite{bailishotos}, and
found that 25 out of every 1000 rows in the database violate at least
one integrity constraint. Bailis \emph{et al.}~\cite{bailisferal} rely
on Rails' \emph{uniqueness validation} to maintain uniqueness of
records while serving Linkbench's~\cite{linkbench} insertion workload
(6400 records distributed over 1000 keys; 64 concurrent clients), and
report discovering more than 10 duplicate records.  Rails relies on
database transactions to validate uniqueness during insertions, which
is sensible if transactions are serializable, but incorrect under the
weak isolation level used in the experiments. The same study has found
that 13\% of all invariants among 67 open source Ruby-on-Rails
applications are at risk of being violated due to weak
isolation. Indeed, incidents of safety violations due to executing
applications in a weakly-isolated environment have been reported on
web services in production~\cite{starbucksbug, scimedbug}, including
in safety-critical applications such as bitcoin
exchanges~\cite{poloniexbug, bitcoinbug}. While enforcing
serializability for all transactions would be sufficient to avoid
these errors and anomalies, it would likely be an overly conservative
strategy; indeed, 75\% of the invariants studied in~\cite{bailisferal}
were shown to be preserved under some form of weak isolation.  When to
use weak isolation, and in what form, is therefore a prominent
question facing all database programmers.\footnote{This position has
  been echoed by leading database researchers who lament the lack of
  better understanding of this problem.  See: {\tt
    http://www.bailis.org/blog/understanding-weak-isolation-is-a-serious-problem}.}

@@

In this paper, we propose a program logic for weakly-isolated
transactions along with automated verification support that realizes
this goal.  We develop a set of syntax-directed compositional proof
rules that allow developers to build correctness proofs for
transactional programs in the presence of a weakly-isolated
concurrency control mechanism.  Realizing that the proof burden to use
these rules would likely discourage applications programmers, we
present an inference procedure based on a state transformer semantics
that allows us to automatically verify the weakest isolation level for
a transaction that ensures its consistency requirements are
maintained.  The key to inference is our formulation of database state
(represented as sets of tuples) as a monad, and in which database
computations are interpreted as transformers over these sets.  This
interpretation leads to an encoding of database computations amenable
for verification by off-the-shelf SMT solvers.

We have realized these ideas as an embedded DSL within OCaml, that
supports common relational SQL operations (updates, selects, inserts,
deletes, joins, etc.).  Experimental results on real-world benchmarks
demonstrate the feasibility and value of automated verification for
weakly isolated transactions, an important advance in improving the
safety of realistic database computations. The paper makes the
following contributions:
\vspace*{-6pt}
\begin{enumerate}
\item An embedded DSL in OCaml that treats SQL-based relational
  database operations as monadic computations over an abstract
  database state.
 \item An operational semantics and compositional proof system for
   this language capable of relating high-level application invariants
   to database state, parameterized by a weak isolation semantics that
   selectively exposes the visibility of these operations to other
   transactions.
 \item An inference algorithm capable of discovering the weakest
   isolation level that is sound with respect to a transaction's
   high-level consistency requirements.  The algorithm interprets
   database operations as predicate state transformers leadig to a
   natural encoding amenable for automated verification.
 \item A realistic implementation along with results on real-world
   database benchmarks that justify our approach and the utility of
   our proof methodology and inference procedure.
\end{enumerate}
\noindent Our results provide the first (to the best of our knowledge)
automated mechanism to verify the safety of high-level consistency
conditions for weakly-isolated transactions, thereby allowing such
transactions to enjoy the same rigorous reasoning capabilities as
their strongly-isolated (serializable) counterparts.

### 2017-07-04 opsem.tex

The small-step semantics is stratified into a transaction-local reduction
relation, and a top-level reduction relation. Transaction-local
relation ($\stg \vdash (c,\stl) \stepsto (c',\stl')$) defines a
small-step reduction for a command inside a transaction, when the
database state is $\stg$; the command $c$ reduces to $c'$, while
updating the transaction-local database $\stl$ to $\stl'$. The
definition assumes a meta function $\eval$ that evaluates expressions
with no free variables to values. The reduction relation for SQL
statements is defined straightforwardly.  \C{INSERT} adds a new record
to $\stl$ after adding a unique identifier. \C{DELETE} finds the records in $\stg$ that match
the search criteria defined by its boolean function argument, and adds
the records to $\stl$ after marking them for deletion. \C{SELECT}
bounds the name introduced by \C{LET} to the set of records from
$\stg$ that match the search criteria, and then executes the bound
command $c$. \C{UPDATE} uses its first function argument to compute
the updated version of the records that match the search criteria
defined by its second function argument. Updated records are added to
$\stl$.

@@

The top-level reduction relation defines the small-step semantics of
transactions, and their parallel composition. A transaction always
comes tagged with an isolation specification $\I$, which has two
components $\I_e$ and $\I_c$, the dictate the timing and nature of
interferences that the transaction can witness, during its execution
($\I_e$), and when it is about to commit ($\I_c$).  Formally, $\I_e$
and $\I_c$ are predicates over by the (current) transaction-local
database state ($\stl$), the state ($\stg$) of the global database
when the transaction last took a step, and the current state ($\stg'$)
of the global database.  Intuitively, $\stg'\neq\stg$ indicates an
interference from another concurrent transaction, and the predicates
$\I_e$ and $\I_c$ decide if this interference is allowed or not,
taking into account the local database state ($\stl$). For instance,
as described in (Sec.~\ref{sec:motivation}), an RR transaction on
PostgreSQL defines $\I$ as following:
\begin{smathpar}
\begin{array}{lcl}
\I_e\,\,(\stl,\stg,\stg') & = & \stg' = \stg\\
\I_c\,\,(\stl,\stg,\stg') & = & \forall(r\in\stl)(r'\in\stg).~ r'.\idf = r.\idf \Rightarrow r'\in\stg'.
\end{array}
\end{smathpar}
For brevity in presentation, we use $\I$ instead of $\I_e$ and $\I_c$
whenever the discussion applies to both $\I_e$ and $\I_c$.

The reduction of a $\ctxn{i}{\I}{c}$ begins by first converting it to
its run-time form $\ctxnr{i}{\I,\stl,\stg}{c}$, where $\stl =
\emptyset$, and $\stg$ the current (global) database.  Rule
\rulelabel{E-Txn} reduces $\ctxnr{i}{\I,\stl,\stg}{c}$ under a
database state ($\stg'$), only if the execution-time isolation
specification ($\I_e$) allows the interference between $\stg$ and
$\stg'$. Rule \rulelabel{E-Commit} commits the transaction
$\ctxnr{i}{\I,\stl,\stg}{c}$ by flushing its uncommitted records to
the database. This is done only if the interference between $\stg$ and
$\stg'$ is allowed at the commit point by the isolation specification
($\I_c$).  The distinction between $\I_e$ and $\I_c$ allows us to
model snapshot semantics of realistic isolation levels that isolate a
transaction from interference during its execution, but exposes
interference at the commit point.

>>>>

The small-step semantics is stratified into a transaction-local reduction
relation, and a top-level reduction relation. The transaction-local
relation ($\stg \vdash (c,\stl) \stepsto (c',\stl')$) defines a
small-step reduction for a command inside a transaction, when the
database state is $\stg$; the command $c$ reduces to $c'$, while
updating the transaction-local database $\stl$ to $\stl'$. The
definition assumes a meta function $\eval$ that evaluates expressions
with no free variables to values. The reduction relation for SQL
statements is defined straightforwardly.  \C{INSERT} adds a new record
to $\stl$ after adding a unique identifier. \C{DELETE} finds the records in $\stg$ that match
the search criteria defined by its boolean function argument, and adds
the records to $\stl$ after marking them for deletion. \C{SELECT}
bounds the name introduced by \C{LET} to the set of records from
$\stg$ that match the search criteria, and then executes the bound
command $c$. \C{UPDATE} uses its first function argument to compute
the updated version of the records that match the search criteria
defined by its second function argument. Updated records are added to
$\stl$.

@@

The top-level reduction relation defines the small-step semantics of
transactions, and their parallel composition. A transaction always
comes tagged with an isolation specification $\I$, which has two
components $\I_e$ and $\I_c$, that dictate the timing and nature of
interferences that the transaction can witness, during its execution
($\I_e$), and when it is about to commit ($\I_c$).  Formally, $\I_e$
and $\I_c$ are predicates over the (current) transaction-local
database state ($\stl$), the state ($\stg$) of the global database
when the transaction last took a step, and the current state ($\stg'$)
of the global database.  Intuitively, $\stg'\neq\stg$ indicates an
interference from another concurrent transaction, and the predicates
$\I_e$ and $\I_c$ decide if this interference is allowed or not,
taking into account the local database state ($\stl$). For instance,
as described in (Sec.~\ref{sec:motivation}), an RR transaction on
PostgreSQL defines $\I$ as following:
\begin{smathpar}
\begin{array}{lcl}
\I_e\,\,(\stl,\stg,\stg') & = & \stg' = \stg\\
\I_c\,\,(\stl,\stg,\stg') & = & \forall(r\in\stl)(r'\in\stg).~ r'.\idf = r.\idf \Rightarrow r'\in\stg'.
\end{array}
\end{smathpar}
For brevity in presentation, we use $\I$ instead of $\I_e$ and $\I_c$
when its destructed form is not required.

The reduction of a $\ctxn{i}{\I}{c}$ begins by first converting it to
its run-time form $\ctxnr{i}{\I,\stl,\stg}{c}$, where $\stl =
\emptyset$, and $\stg$ the current (global) database.  Rule
\rulelabel{E-Txn} reduces $\ctxnr{i}{\I,\stl,\stg}{c}$ under a
database state ($\stg'$), only if the execution-time isolation
specification ($\I_e$) allows the interference between $\stg$ and
$\stg'$. Rule \rulelabel{E-Commit} commits the transaction
$\ctxnr{i}{\I,\stl,\stg}{c}$ by flushing its uncommitted records to
the database. This is done only if the interference between $\stg$ and
$\stg'$ is allowed at the commit point by the isolation specification
($\I_c$).  The distinction between $\I_e$ and $\I_c$ allows us to
model the snapshot semantics of realistic isolation levels that
isolate a transaction from interference during its execution, but
expose interferences at the commit point.
